---
title: 'mefa4 Design Decisions and Performance'
author: "Peter Solymos"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  pdf_document:
    keep_tex: no
    toc: true
    toc_depth: 3
    number_sections: true
  word_document:
    toc: true
    toc_depth: 3
documentclass: article
fontsize: 11pt
classoption: letterpaper
---

> **mefa4** is a reimplementation of the S3 object classes 
> found in the **mefa** R package.
> The new S4 class `"Mefa"` has all the consistency checks that S3 classes
> can not have, and most importantly, it stores the 
> cross-tabuted results as a compact
> sparse matrix (S4 object class `"dgCMatrix"` of the **Matrix** package).
> The use of sparse matrices speed up computations,
> and reduces object sizes considerably. 
> This vignette introduces the main functions,
> classes and methods of the package **mefa4**.

# Introduction

The aim of the **mefa** and **mefa4** packages are to 
help in storing cross tabulated
ecological data tables (community data) together with 
attributes for rows (samples) and columns
(species, taxa). This allows that one can easily subset 
the relational data object without
separately manipulating 2--3 pieces of R objects. 
By doing so, the chances of errors are reduced.

As ecological data sets are increasing in size, it is 
necessary to find more efficient ways of
data storage and manipulation. To this end, it was in 
the air for some time to redesign
the `mefa` package and take advantages of sparse matrices 
from the **Matrix** package.
This is done at the costs of some old functionalities not being 
available for S4 classes at the time being.
Here I give an overview so the user can decide how to use the 
parallel availability of old S3 and newer S4 classes.

# Comparisons

## S3 and S4 classes

The S3 classes defined in **mefa** were `stcs` and `mefa`.
`stcs` is a data frame with several attributes:

```{r}
library(mefa)
x <- data.frame(
    sample = paste("Sample", c(1,1,2,2,3,4), sep="."),
    species = c(paste("Species", c(1,1,1,2,3), sep="."),  "zero.pseudo"),
    count = c(1,2,10,3,4,0),
    segment = letters[c(6,13,6,13,6,6)],
    stringsAsFactors = TRUE)
s <- stcs(x)
attributes(s)
```

These attributes ensure that the cross-tab made by the function 
`mefa()` creates a proper
cross-tab by eliminating the column that is only a placeholder 
for empty samples, etc.:

```{r}
samp <- data.frame(samples=levels(x$sample), var1=1:2,
    stringsAsFactors = TRUE)
taxa <- data.frame(specnames=levels(x$species), var2=c("b","a"),
    stringsAsFactors = TRUE)
rownames(samp) <- samp$samples
rownames(taxa) <- taxa$specnames
(m <- mefa(s, samp, taxa))
m$xtab
```

The `stcs` step is almost redundant, and inefficient 
relative to the `stats::xtabs`
function with `sparse = TRUE`. This function is 
adapted to some extent, so it can subset
the cross-tabulated results before returning the 
value (`rdrop` and `cdrop` arguments,
that is available as the `Xtab` function in the 
**mefa4** package). This takes a formula,
and can be applied directly on a data frame. 
The formula can have a left-hand side, or the left-hand side can be 
missing. The right-hand side can contain 2--3 factors, 
and the result will be a sparse matrix
or a list of sparse matrices, respectively:

```{r}
library(mefa4)
x0  <- Xtab(~ sample + species, x)
x1  <- Xtab(count ~ sample + species, x)
x11 <- Xtab(count ~ sample + species + segment, x)
```

Dropping some rows/columns can be done in several ways.
A logical statement implies that all empty rows/columns are dropped,
but indices (numeric or character) can also be used:

```{r}
x2  <- Xtab(count ~ sample + species, x, cdrop=FALSE, rdrop=TRUE)
x21 <- Xtab(count ~ sample + species, x, cdrop=TRUE, rdrop=FALSE)
(x22 <- Xtab(count ~ sample + species, x, cdrop="zero.pseudo"))
```

The results here are sparse matrices in compact mode, this means that
redundant indices are only kept once, so it is more compact than
a long formatted database representation stored in an `stcs`
object or in the original data frame, or a triplet representation 
of a sparse matrix. See vignettes in the 
**Matrix** package for more details on S4 sparse matrix classes.

The S4 class `"Mefa"` is defined in the **mefa4** package.
It can be created by the `Mefa()` function, and the result has 4 slots:

```{r}
(x3 <- Mefa(x1, samp, taxa))
```

The `xtab` slot stores the cross-tab in sparse matrix format.
The `samp` slot stores the row attributes for `xtab`
as data frame or can be `NULL`.
The `taxa` slot stores the column attributes for `xtab` 
as data frame or can be `NULL`. Validity checks are done to ensure
proper object classes to be used and matching dimnames.
The option that a column in the attribute tables can be specified
to find matching names is not available in the new implementation.
Corresponding rownames of the data frames has to match dimnames of `xtab`.
The `join` slot can be `"left"` (all rows/columns in the
`xtab` are kept, matching attributes are selected, non-matching
attributes are excluded, and missing attributes are filled up with `NA`)
or `"inner"` (only the intersection of corresponding dimnames are used to
form the return value).

The call in `Mefa()` can take any matrix or sparse matrix as argument,
but it will be stored in a sparse mode. Here we use a matrix as input,
and `samp` has missing values (`"left"` join is used by default):
```{r}
(x4 <- Mefa(as.matrix(x1), samp[1:2,]))
```
The effect of `"inner"` join is as follows:
```{r}
(x5 <- Mefa(x2, samp, taxa, join="inner"))
(x51 <- Mefa(x2, samp[1:2,], taxa, join="inner"))
```
A `"Mefa"` object with only `xtab` can also be defined:
```{r}
(x6 <- Mefa(x1))
```

The equivalent of the `melt` method of the **mefa** package is
the `Melt` function in **mefa4**. It can be used to reverse
the side effects of the cross-tabulation, thus making a data frame
from a matrix, sparse matrix, list of sparse matrices, 
a `mefa` or a `Mefa` object:

```{r}
Melt(x1)
Melt(x11)
```

The structure of the S3 and S4 classes are very similar, and even the
accessor methods (`xtab()`, `samp()`, `taxa()`, `segm()`)
work properly on both types. The S4 class does not have a slot for
a call, and there is no `segm` element/slot either. This means that
a `"Mefa"` object cannot have 3 dimensions, only 2. `Xtab` can create
3-dimensional sparse array-like objects (list of sparse matrices of the same dimensions),
but there is no formal S4 class that can handle sparse matrix lists as part of 
a `"Mefa"` object.
The `as.mefa` method can convert such a list of sparse matrices into an S3
`"mefa"` object with segments.

## Coercions back and forth

Coercion methods are defined in both the **mefa** and **mefa4** packages
to ensure that S3 and S4 objects are interchangeable:

```{r}
as.stcs(x1)
as.mefa(x1)
as.stcs(x3)
a <- as.mefa(x3)
xtab(a)
samp(a)
taxa(a)
segm(a)
segm(x3)
as.Mefa(a)
as.Xtab(a)
s <- melt(a)
as.Xtab(s)
as.Mefa(s)
melt(x1)
melt(x3)
```

## Subsetting and replacement

Accessing and replacing parts of the `"Mefa"` object
is conveniently done by methods `xtab`, `samp`,
and `taxa` (the `segm` S3 method only returns the
`xtab` slot of an S4 `"Mefa"` object):

```{r}
xtab(x3)
x1[3,1] <- 999
xtab(x3) <- x1
xtab(x3)
```

Attribute tables can be set to `NULL`, or replaced:

```{r}
samp(x3)
samp(x3) <- NULL
samp(x3)
samp(x3) <- samp[1:3,]
samp(x3)
```

```{r}
taxa(x3)
taxa(x3) <- NULL
taxa(x3)
taxa(x3) <- taxa[1:3,]
taxa(x3)
```

Replacing parts of these attribute tables can be done as

```{r}
samp(x3)[1,]
samp(x3)[1,2] <- 3
samp(x3)[1,]
```

Subsetting the whole `"Mefa"` object is done via the `[`
method:

```{r}
x3[3:2, 1:2]
x3[3:2, ]
x3[ ,1:2]
```

# Methods for S4 classes

Simple methods are provided for convenience:

```{r}
dim(x5)
dimnames(x5)
dn <- list(paste("S", 1:dim(x5)[1], sep=""),
    paste("SPP", 1:dim(x5)[2], sep=""))
dimnames(x5) <- dn
dimnames(x5)[[1]] <- paste("S", 1:dim(x5)[1], sep="_")
dimnames(x5)[[2]] <- paste("SPP", 1:dim(x5)[2], sep="_")
t(x5)
```

# Utilities

## Grouping rows and columns

The `aggregate` method was defined for 
S3 `mefa` objects. Its equivalent (although it cannot
sum the cells simultaneously for rows and columns, but
it was done in 2 subsequent steps anyway) is the
`groupSums` method. The `MARGIN` argument
indicates if rows (`MARGIN = 1`) or columns
(`MARGIN = 2`) are to be added together:

```{r}
groupSums(as.matrix(x2), 1, c(1,1,2))
groupSums(as.matrix(x2), 2, c(1,1,2,2))
groupSums(x2, 1, c(1,1,2))
groupSums(x2, 2, c(1,1,2,2))
groupSums(x5, 1, c(1,1,2))
groupSums(x5, 2, c(1,1,2,2))
```
A simple extension of this is the `groupMeans`
method:

```{r}
groupMeans(as.matrix(x2), 1, c(1,1,2))
groupMeans(as.matrix(x2), 2, c(1,1,2,2))
groupMeans(x2, 1, c(1,1,2))
groupMeans(x2, 2, c(1,1,2,2))
groupMeans(x5, 1, c(1,1,2))
groupMeans(x5, 2, c(1,1,2,2))
```

## Combining objects

`mbind` can be used to combine 2 matrices (dense or sparse).
The 2 input objects are combined in a left join manner, which means 
that all the elements in the first object are retained, 
and only non-overlapping elements in the second object are used.
Elements of the returning object that are not part of either objects (outer set)
are filled up with value provided as `fill` argument.

```{r}
x=matrix(1:4,2,2)
rownames(x) <- c("a", "b")
colnames(x) <- c("A", "B")
y=matrix(11:14,2,2)
rownames(y) <- c("b", "c")
colnames(y) <- c("B", "C")
mbind(x, y)
mbind(x, y, fill=0)
mbind(as(x, "sparseMatrix"), as(y, "sparseMatrix"))
```

`"Mefa"` objects can be combined in a similar way, where
attribute tables are combined in a left join fashion
(S3 `"mefa"` objects have to be coerced
by the `as.Mefa` method beforehand -- this is so
because the S3 class does not allow `NA` values in
`$xtab`, and it is safer to avoid unnecessary complications):

```{r}
sampx <- data.frame(x1=1:2, x2=2:1,
    stringsAsFactors = TRUE)
rownames(sampx) <- rownames(x)
sampy <- data.frame(x1=3:4, x3=10:11,
    stringsAsFactors = TRUE)
rownames(sampy) <- rownames(y)
taxay <- data.frame(x1=1:2, x2=2:1,
    stringsAsFactors = TRUE)
rownames(taxay) <- colnames(y)
taxax <- NULL
mbind(Mefa(x, sampx), Mefa(y, sampy, taxay))
```

# Performance comparisons

We compare the performance of the **mefa** and **mefa4**
packages. We are using a long formatted raw data file from the Alberta
Biodiversity Monitoring Institute database (available at [https://www.abmi.ca](https://www.abmi.ca)):

```{r}
library(mefa)
library(mefa4)
data(abmibirds)
```

This is the processing with **mefa** and S3 object classes 
(we are storing the results and processing times):

```{r}
b3 <- abmibirds
b3 <- b3[!(b3$Scientific.Name %in% c("VNA", "DNC", "PNA")),]
levels(b3$Scientific.Name)[levels(b3$Scientific.Name) 
    %in% c("NONE", "SNI")] <- "zero.pseudo"
b3$Counts <- ifelse(b3$Scientific.Name == "zero.pseudo", 0, 1)
b3$Label <- with(b3, paste(ABMI.Site, Year, 
    Point.Count.Station, sep="_"))
x3 <- b3[!duplicated(b3$Label), c("Label", 
    "ABMI.Site", "Year", "Field.Date", 
    "Point.Count.Station", "Wind.Conditions", "Precipitation")]
rownames(x3) <- x3$Label
z3 <- b3[!duplicated(b3$Scientific.Name), c("Common.Name",
    "Scientific.Name", "Taxonomic.Resolution", 
    "Unique.Taxonomic.Identification.Number")]
rownames(z3) <- z3$Scientific.Name
z3 <- z3[z3$Scientific.Name != "zero.pseudo",]
t31 <- system.time(s3 <- suppressWarnings(stcs(b3[,
    c("Label","Scientific.Name","Counts")])))
t32 <- system.time(m30 <- mefa(s3))
t33 <- system.time(m31 <- mefa(s3, x3, z3))
y30 <- m30$xtab
t34 <- system.time(m32 <- mefa(y30, x3, z3))
m32
```

The equivalent processing with **mefa4** and S4 object classes:

```{r}
b4 <- abmibirds
b4$Label <- with(b4, paste(ABMI.Site, Year, 
    Point.Count.Station, sep="_"))
x4 <- b4[!duplicated(b4$Label), c("Label", "ABMI.Site", 
    "Year", "Field.Date", "Point.Count.Station",
    "Wind.Conditions", "Precipitation")]
rownames(x4) <- x4$Label
z4 <- b4[!duplicated(b4$Scientific.Name), c("Common.Name",
    "Scientific.Name", "Taxonomic.Resolution", 
    "Unique.Taxonomic.Identification.Number")]
rownames(z4) <- z4$Scientific.Name
t41 <- system.time(s4 <- Xtab(~ Label + Scientific.Name, 
    b4, cdrop = c("NONE", "SNI"), 
    subset = !(b4$Scientific.Name %in% c("VNA", "DNC", "PNA")), 
    drop.unused.levels = TRUE))
t42 <- system.time(m40 <- Mefa(s4))
t43 <- system.time(m41 <- Mefa(s4, x4, z4))
y40 <- as.matrix(m40@xtab)
t44 <- system.time(m42 <- Mefa(y40, x4, z4))
m42
sum(m42@xtab)
```

Let us compare object sizes and processing times, stars indicate
similar S3 (`*=3`) and S4 (`*=4`) objects:

```{r}
res <- cbind("SIZE, *=3"=c("b*"=object.size(b3),
    "s*"=object.size(s3),
    "y*0"=object.size(y30),
    "m*0"=object.size(m30),
    "m*1"=object.size(m31),
    "m*2"=object.size(m32)),
"SIZE, *=4"=c("b*"=object.size(b4),
    "s*"=object.size(s4),
    "y*0"=object.size(y40),
    "m*0"=object.size(m40),
    "m*1"=object.size(m41),
    "m*2"=object.size(m42)),
"TIME, *=3"=c("b*"=NA,
    "s*"=t31[3],
    "y*0"=NA,
    "m*0"=t32[3],
    "m*1"=t33[3],
    "m*2"=t34[3]),
"TIME, *=4"=c("b*"=NA,
    "s*"=t41[3],
    "y*0"=NA,
    "m*0"=t42[3],
    "m*1"=t43[3],
    "m*2"=t44[3]))
(res <- cbind(res, "SIZE"=res[,2]/res[,1], "TIME"=res[,4]/res[,3]))
```

The compressed sparse matrix representation is `r round(100*res[2,5], 1)`% of the `stcs` object in size.
`"Mefa"` object sizes are maximum of `r round(100*max(res[5:6,5]),1)`% of their S3 representatives.
Processing time speed-up is enormous with sparse matrices (`r round(100*mean(res[4:5,6]),1)`%), and still
quite high by standard matrices (`r round(100*res[6,6],1)`%).

Check that objects are the same:

```{r}
stopifnot(identical(dim(y30), dim(y40)))
stopifnot(identical(setdiff(rownames(y30), rownames(y40)), character(0)))
stopifnot(identical(setdiff(rownames(y40), rownames(y30)), character(0)))
stopifnot(identical(setdiff(colnames(y30), colnames(y40)), character(0)))
stopifnot(identical(setdiff(colnames(y40), colnames(y30)), character(0)))
```

The aggregation also improved quite a bit with sparse matrices:

```{r}
system.time(xx3 <- aggregate(m31, "ABMI.Site"))
system.time(xx4 <- groupSums(m41, 1, m41@samp$ABMI.Site))
```

# Conclusions

The redesign of the old S3 classes into S4 ones resulted in large savings in computing time
and object sizes. Old features are still available due to the free conversion
between the two implementations.
